Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-11-14 paper/appendix.tex
fixes

Consequently, assuming that $F$ is full rank on $Img(XE)$, B2B yields the same solutions to equations \ref{eq:model} and \ref{model_noisy_measure}.

@@

\subsection{Recovering E}

@@

First, given known signal-to-noise ratio,the threshold above which a feature should considered to be causal can be derived analytically. Indeed, Equation $\ref:diagk$ implies that the $k$ first diagonal elements of $\hat H$ are bounded:  % we assume that $F$ has full rank over $Img(E)$.

@@

Second, $\text{diag{\hat H}}$ can be binarized with the Sonquist-Morgan criterion \citep{sonquist_morgan}, a non-parametric clustering procedure separating small and large values in a given set. This procedure maximizes the ratio of inter-group variance while minimizing the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values. Let $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, Sonquist-Morgan criterion maximizes \citep{kass75}:
\begin{equation}
  \frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}
\end{equation}

@@

To test whether each of the methods robustly scales to an increasingly large number of potential causes $X$, we enhanced the four ad-hoc features (word length, word frequency, word function, dummy variable) with another ten features. These additional features corresponds to the first dimensions of word embedding as provided by Spacy \citep{spacy2}. The results shown in Figure \ref{fig:embbedings}, show that the feature importance of ad-hoc features as derived by B2B remain unchanged (and are actually improved).

>>>>

Consequently, assuming that $F$ is full rank on $Img(XE)$, B2B yields the same solutions to equations \ref{eq:model} and \ref{eq:model_noisy_measure}.

@@

\subsection{Recovering E}
\label{recovering}

@@

First, given known signal-to-noise ratio,the threshold above which a feature should considered to be causal can be derived analytically. Indeed, Equation $\ref{eq:diagk}$ implies that the $k$ first diagonal elements of $\hat H$ are bounded:  % we assume that $F$ has full rank over $Img(E)$.

@@

Second, $\text{diag}(\hat H)$ can be binarized with the Sonquist-Morgan criterion \citep{sonquist_morgan}, a non-parametric clustering procedure separating small and large values in a given set. This procedure maximizes the ratio of inter-group variance while minimizing the intra-group variance, over all possible splits of the diagonal into $p$ largest values and $d_x-p$ smallest values. Let $m_0$ and $m_1$ be the average values of the two clusters, $p$ and $d_x-p$ their size, and $v$ the total variance of the sample, Sonquist-Morgan criterion maximizes \citep{kass75}:
\begin{equation}
  \frac{p(d_x-p)}{d_x} \frac{(m_1 - m_0)^2}{v}
\end{equation}

@@

To test whether each of the methods robustly scales to an increasingly large number of potential causes $X$, we enhanced the four ad-hoc features (word length, word frequency, word function, dummy variable) with another ten features. These additional features corresponds to the first dimensions of word embedding as provided by Spacy \citep{spacy2}. The results shown in Figure \ref{fig:embeddings}, show that the feature importance of ad-hoc features as derived by B2B remain unchanged and are actually improved.

### 2019-11-14 paper/introduction.tex
fixes

The present paper focuses on the restricted issue of disentangling the causal influence of linearly correlated predictors ($X$) onto noisy multivariate observations ($Y$). The present approach thus differs from other causal discovery algorithms based on temporal-delays and/or nonlinear interaction in systems where the directionality of causation (from X to Y or vice versa) is unknown (e.g. \citep{peters2017elements, granger1969investigating, janzing2013quantifying, scholkopf2016modeling}.

>>>>

The present paper focuses on the restricted issue of disentangling the causal influence of linearly correlated predictors ($X$) onto noisy multivariate observations ($Y$). The present approach thus differs from other causal discovery algorithms based on temporal-delays and/or nonlinear interactions in systems where the directionality of causation (from X to Y or vice versa) is unknown (e.g. \citep{peters2017elements, granger1969investigating, janzing2013quantifying, scholkopf2016modeling}.

### 2019-11-14 paper/paper.tex
fixes

Assuming, without loss of generality, that the active features in $E$ are the $k \in \mathbb{Z}: k \in [0, d_x]$ first features, and rewriting $X=(X_1,X_2)$ and $N=(N_1,N_2)$ ($X_1$ and $N_1$ containing the $k$ first features), we have ($\Sigma_{A B}$ denoting the covariance of $A$ and $B$)

% $$
\begin{equation}
  X^\top X = \left(\begin{array}{lccl}\Sigma_{X_1 X_1} & \Sigma_{X_1 X_2} \\ \Sigma_{X_1 X_2} & \Sigma_{X_2 X_2}\end{array}\right),\qquad\qquad N^\top N = \left(\begin{array}{lccl}\Sigma_{N_1 N_1} & \Sigma_{N_1 N_2} \\ \Sigma_{N_1 N_2} & \Sigma_{N_2 N_2}\end{array}\right),
\end{equation}
% $$
% and
\begin{equation}
  \hat H = \left(\begin{array}{cc}(\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1} & (\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_2} \\0 & 0\end{array}\right),
\end{equation}
\begin{equation}
  \text{diag}_k (\hat H) = \text{diag}((\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1}) = \text{diag}((I+\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1})^{-1}).
  \label{eq:diagk}
\end{equation}
%
% $$
% $$
% $$
% $$
%
In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so $\text{diag}_k(\hat H)=I$, and $$\text{diag}(\hat H) = \text{diag}(E)$$ Therefore, we recover $E$ from $\hat H$.

Otherwise, with noise, the causal factors of $E$ correspond to the positive elements of $\test{diag}(\hatH)$. The methods to recover them are presented in Appendix XXXX$$$

@@

We perform two experiments to evaluate B2B: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography (MEG) dataset.
%
We use scikit-learn's PLS and RidgeCV \citep{sklearn} as well as pyrcca's regularized canonical component analysis (RegCCA, \citep{bilenko2016pyrcca}) objects to compare B2B against the standard baselines.

>>>>

Assuming, without loss of generality, that the active features in $E$ are the $k \in \mathbb{Z}: k \in [0, d_x]$ first features, and rewriting $X=(X_1,X_2)$ and $N=(N_1,N_2)$ ($X_1$ and $N_1$ containing the $k$ first features), we have:

% $$
\begin{equation}
  X^\top X = \left(\begin{array}{lccl}\Sigma_{X_1 X_1} & \Sigma_{X_1 X_2} \\ \Sigma_{X_1 X_2} & \Sigma_{X_2 X_2}\end{array}\right),\qquad\qquad N^\top N = \left(\begin{array}{lccl}\Sigma_{N_1 N_1} & \Sigma_{N_1 N_2} \\ \Sigma_{N_1 N_2} & \Sigma_{N_2 N_2}\end{array}\right),
\end{equation}

where $\Sigma_{A B}$ is the covariance of $A$ and $B$, and:
% and
\begin{equation}
  \hat H = \left(\begin{array}{cc}(\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1} & (\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_2} \\0 & 0\end{array}\right)
\end{equation}
\begin{equation}
  \text{diag}_k (\hat H) = \text{diag}((\Sigma_{X_1 X_1}+\Sigma_{N_1 N_1})^{-1}\Sigma_{X_1 X_1}) = \text{diag}((I+\Sigma_{X_1 X_1}^{-1}\Sigma_{N_1 N_1})^{-1})
  \label{eq:diagk}
\end{equation}
%
% $$
% $$
% $$
% $$
%
In the absence of noise, we have $\Sigma_{N_1 N_1}=0$, and so $\text{diag}_k(\hat H)=I$, and $$\text{diag}(\hat H) = \text{diag}(E)$$ Therefore, we recover $E$ from $\hat H$.

In the presence of noise, the causal factors of $E$ correspond to the positive elements of $\text{diag}(\hat H)$. The methods to recover them are presented in the Appendix \ref{recovering}.

@@

We perform two sets of experiments to evaluate B2B: one on controlled synthetic data, and a second one on a real, large-scale magneto-encephalography (MEG) dataset.
%
We use scikit-learn's PLS and RidgeCV \citep{sklearn} as well as pyrcca's regularized canonical component analysis (RegCCA, \citep{bilenko2016pyrcca}) objects to compare B2B against the standard baselines.
